# Assignment 4

## Data wrangling

The create_human.R rscipt can be found here: <https://github.com/emselina/IODS-project/blob/master/create_human.R>

## Data analysis

*Load the Boston data from the MASS package. Explore the structure and the dimensions of the data and describe the dataset briefly, assuming the reader has no previous knowledge of it. Details about the Boston dataset can be seen for example here. (0-1 points)*


```{r}
# access the MASS package
library(MASS)

# load the data
data(Boston)

#Explore the structure and the dimensions of the data
dim(Boston)
str(Boston)
```
The Boston data frame has 506 rows and 14 columns.

This dataset contains information collected by the U.S Census Service concerning housing in the area of Boston Mass.

Details of the dataset:
crim : per capita crime rate by town.

zn : proportion of residential land zoned for lots over 25,000 sq.ft.

indus : proportion of non-retail business acres per town.

chas :Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).

nox : nitrogen oxides concentration (parts per 10 million).

rm : average number of rooms per dwelling.

age :proportion of owner-occupied units built prior to 1940.

dis :weighted mean of distances to five Boston employment centres.

rad :index of accessibility to radial highways.

tax :full-value property-tax rate per $10,000.

ptratio :pupil-teacher ratio by town.

black : 1000(Bk - 0.63)^21000(Bk−0.63) 2 where Bk is the proportion of blacks by town.

lstat : lower status of the population (percent).

medv : median value of owner-occupied homes in $1000s.

*3. Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. (0-2 points)*

```{r}
summary(Boston)
```
```{r}
# access the GGally and ggplot2 libraries
library(GGally)
library(ggplot2)
pairs(Boston)
```
I think this looks ugly and I can not see anything, but maybe it gives some kind of overview. Lets try something else. 


```{r}
p1 <- ggplot(Boston) + 
  geom_histogram(aes(x = crim), fill = "#E69F00") +
labs(title="per capita crime rate by town")

p2 <- ggplot(Boston) + 
    geom_histogram(aes(x = zn), fill = "#56B4E9") +
  labs(title="proportion of residential land zoned for lots over 25,000 sq.ft.")

p3 <- ggplot(Boston) + 
    geom_histogram(aes(x = indus), fill = "#009E73")+
  labs(title="proportion of non-retail business acres per town")

p4 <- ggplot(Boston) + 
    geom_histogram(aes(x = chas), fill = "#F0E442")+
  labs(title="Charles River dummy variable")

p5 <- ggplot(Boston) + 
    geom_histogram(aes(x = nox), fill = "#0072B2")+
  labs(title="nitrogen oxides concentration")

p6 <- ggplot(Boston) + 
  geom_histogram(aes(x = rm), fill = "#D55E00")+
  labs(title="average number of rooms per dwelling.")

p7 <- ggplot(Boston) + 
  geom_histogram(aes(x = age), fill = "wheat1")+
  labs(title="proportion of owner-occupied units built prior to 1940")


p8 <- ggplot(Boston) + 
  geom_histogram(aes(x = dis), fill = "green")+
  labs(title="weighted mean of distances to five Boston employment centres")

p9 <- ggplot(Boston) + 
  geom_histogram(aes(x = rad ), fill = "mediumorchid2")+
  labs(title="index of accessibility to radial highways")

p10 <- ggplot(Boston) + 
  geom_histogram(aes(x = tax  ), fill = "grey")+
  labs(title="full-value property-tax rate per $10,00")

p11 <- ggplot(Boston) + 
  geom_histogram(aes(x = ptratio  ), fill = "pink4")+
  labs(title="pupil-teacher ratio by town")

p12 <- ggplot(Boston) + 
  geom_histogram(aes(x = black  ), fill = "black")+
  labs(title="1000(Bk−0.63)2 where Bk is the proportion of blacks by town.")

p13 <- ggplot(Boston) + 
  geom_histogram(aes(x = lstat  ), fill = "seagreen2")+
  labs(title="lower status of the population")


p14 <- ggplot(Boston) + 
  geom_histogram(aes(x = medv  ), fill = "coral")+
  labs(title="median value of owner-occupied homes in $1000s.")



```



```{r}

library(gridExtra)

gridExtra::grid.arrange(p1, p2, p3, p4, nrow=2)
```


```{r}

library(gridExtra)

gridExtra::grid.arrange(p5, p6, p7, p8, nrow=2)
```



```{r}

library(gridExtra)

gridExtra::grid.arrange(p9, p10, p11, nrow=2)
```










```{r}

library(gridExtra)

gridExtra::grid.arrange(p12, p13, p14, nrow=2)
```




```{r}
# calculate the correlation matrix and round it
cor_matrix <- cor(Boston) %>% round(digits = 2)

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
library(corrplot)
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d",  tl.cex = 0.6)
```
This correlation plot tells the correlation between variables, the bigger and more colourfull the circles are the more the is correlation between the variables. The red color indicates negative correlation and blue color positive correlation. The highest positive correlation is between tax and rad variables (0.91). The biggest negative correlation is between nox and dis variables (-0.77). Second highest negative correlation is between dis and age (0.75). 

*4. Standardize the dataset and print out summaries of the scaled data. How did the variables change? Create a categorical variable of the crime rate in the Boston dataset (from the scaled crime rate). Use the quantiles as the break points in the categorical variable. Drop the old crime rate variable from the dataset. Divide the dataset to train and test sets, so that 80% of the data belongs to the train set. (0-2 points)*


```{r}
# center and standardize variables
boston_scaled <- scale(Boston) 

# summaries of the scaled variables

summary(boston_scaled)
# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)




```

```{r}

boston_scaled$crim <- as.numeric(boston_scaled$crim)
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- c(quantile(boston_scaled$crim))
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels =c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

```

```{r}
library(tidyverse)
boston_scaled <- boston_scaled %>% 
mutate(crime = crime)

# remove the crime variable from test data
boston_scaled <- dplyr::select(boston_scaled, -crim)

```






```{r}

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]



```



*5. Fit the linear discriminant analysis on the train set. Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables. Draw the LDA (bi)plot. (0-3 points)*

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit
```
```{r}
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes = as.numeric(train$crime)


# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 2)

```
The LDA analysis separeted the high crime classes form the low and medium high crime classes. The variable that effects the most is the rad variable = accessibility to radial highways. There is also some separation between low and medium high crime classes and the varibles that mostly affect to this are zn =  proportion of residential land zoned for lots over 25,000 sq.ft. and nox =  nitrogen oxides concentration (parts per 10 million). The LDI 1 separates 95 % of the populatio, LDI 2, 3,5 % and LDI 3 1,5 % of the population. 



*6. Save the crime categories from the test set and then remove the categorical crime variable from the test dataset. Then predict the classes with the LDA model on the test data. Cross tabulate the results with the crime categories from the test set. Comment on the results. (0-3 points)*



```{r}
# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
```


```{r}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes , predicted = lda.pred$class)

```

The model predicted the high crime rate well since all predicted observations are in the high crime category. The model did not predict the low crime category well. 


*7. Reload the Boston dataset and standardize the dataset (we did not do this in the Exercise Set, but you should scale the variables to get comparable distances). Calculate the distances between the observations. Run k-means algorithm on the dataset. Investigate what is the optimal number of clusters and run the algorithm again. Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results. (0-4 points)*


```{r}
# access the MASS package
library(MASS)

# load the data
data(Boston)

# center and standardize variables
boston_scaled <- scale(Boston) 

# euclidean distance matrix
dist_eu <- dist(Boston)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(Boston, method= "manhattan")

# look at the summary of the distances
summary(dist_man)

```

```{r}
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <- kmeans(Boston, centers = 4)

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)

```
Optimal number of clusters is when the total WCSS drops radically so according to the plot about 2 clusters.


```{r}
set.seed(123)

# k-means clustering
km <- kmeans(Boston, centers = 2)

```


```{r}

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)

# plot the Boston dataset with clusters
pairs(Boston[1:7], col = km$cluster)

# plot the Boston dataset with clusters
pairs(Boston[8:14], col = km$cluster)

```
The two clusters are clearly separates in some of the variables. 



*Bonus: Perform k-means on the original Boston data with some reasonable number of clusters (> 2). Remember to standardize the dataset. Then perform LDA using the clusters as target classes. Include all the variables in the Boston data in the LDA model. Visualize the results with a biplot (include arrows representing the relationships of the original variables to the LDA solution). Interpret the results. Which variables are the most influential linear separators for the clusters? (0-2 points to compensate any loss of points from the above exercises)*



```{r}


# load the data
data(Boston)

# center and standardize variables
boston_scaled_2 <- scale(Boston) 

# change the object to data frame
boston_scaled_2 <- as.data.frame(boston_scaled_2)

# k-means clustering
km <- kmeans(boston_scaled_2, centers = 4)

boston_scaled_2 <- boston_scaled_2 %>% 
mutate(clusters = c(km$cluster))

# linear discriminant analysis
lda.km <- lda(clusters ~ ., data = boston_scaled_2)

```



```{r}
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric

classes <- as.numeric(boston_scaled_2$clusters)

# plot the lda results
plot(lda.km, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.km, myscale = 3)
```
The four clusters have separated well. The zn, rad, tax, nox and age and chas are the variables that affect the separation most. 

*Super-Bonus: Run the code below for the (scaled) train data that you used to fit the LDA. The code creates a matrix product, which is a projection of the data points.*


```{r}
model_predictors <- dplyr::select(train, -crime)
# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
```

Next, install and access the plotly package. Create a 3D plot (cool!) of the columns of the matrix product using the code below.


```{r}
#install.packages("plotly")
library(plotly)
```


```{r}
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers')
```

Adjust the code: add argument color as a argument in the plot_ly() function. Set the color to be the crime classes of the train set. Draw another 3D plot where the color is defined by the clusters of the k-means. How do the plots differ? Are there any similarities? (0-3 points to compensate any loss of points from the above exercises)


```{r}
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = train$crime)
```


```{r}
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = km$cluster[ind])
```

The first plot shows shows that the two clusters are clearly separated. In the second plot high crime rate has separated from the other crime levels. In the third plot the fourth cluster is clearly separated from the other clusters just like in plot 2. 



