# Assignment 6. 

## Data wrangling


The data wrangling part and the R scipt created are in the data folder in: <https://github.com/emselina/IODS-project/blob/master/meet_and_repeat.R>


## Data analysis



Note that you must SWAP the data sets! :) It is NOT a simple copy & paste of the MABS book!

*Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)*

In the rats data the groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately) weekly, except in week seven when two recordings were taken) over a 9-week period. The question of most interest is whether the growth profiles of the three groups differ.

```{r}
library(dplyr)

#original rats data in wide format
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep = '\t', header = T)


#RATS data in long format
RATSL <- read.table("data/RATSL.csv", header = T, sep=",")


# Factor variables ID and Group

RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# Glimpse the data

glimpse(RATSL)

#structure of the data
str(RATSL)

#dimensions of the data
dim(RATSL)

#variable names of the data
names(RATSL)

#view the data in table format
#view(RATSL)

#summaries of the variables
summary(RATSL)

```

The Rats data set was wrangled into long format in the data wrangling exercise. Variable Time was added as own one column as numeric values and it presents the day that weight measurement was done. Before in the wide format the time points were each in separate columns. 


```{r}
library(dplyr); library(tidyr);library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group))+
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))+
  scale_y_continuous(name = "Weight (grams)")+
  theme(legend.position = "top")
```

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = Group)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))+
  scale_y_continuous(name = "Weight (grams)")
```



Standardized values will be done by obtaining by subtracting the relevant occasion mean from the original observation and then dividing by the corresponding visit standard deviation

Standardize the values and plot again. 

```{r}
library(dplyr)
library(tidyr)
# Standardise the variable bprs
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdrats = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)
```





Plot with the standardized values:

```{r}
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = stdrats, linetype = Group)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized rats")
```





```{r}
library(dplyr); library(tidyr);library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, group = ID, color = Group)) +
  geom_boxplot()+
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))+
  scale_y_continuous(name = "Weight (grams)")+
  theme(legend.position = "top")
```






Here is presented average (mean) profiles for each treatment group along with some indication of the variation of the observations at each time point, in this case the standard error of mean:


```{r}
# Summary data with mean and standard error of Weight by Group and Time

# Number of subjects (per group):
#n = n() this will calculate the n in each group within the summarise ()

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
    summarise(mean = mean(Weight), n = n(), se = sd((Weight)/sqrt(n))) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)
```


Mean profiles are plotted as line box plots of the observations at each time point in each group. 

```{r, fig.width=6, fig.height=5}


# Plot the mean profiles
library(ggplot2)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="3"), width=0.3) +
  theme(legend.position = "top") +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Applying the Summary Measure Approach


Plotting as boxplot to find possible outliers. 


```{r}

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0)
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)


# Draw a boxplot of the mean versus Group
library(ggplot2)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days 1-60")
```


Some outliers were found in the data in each group. In next step I filtered out outliers that were bigger than mean weight 550 and smaller than mean weight 250. I did not know how to filter out the outlier in the group 3?


```{r}

#filter out outliers
RATSL8S1 <- filter(RATSL8S, mean < 550 & mean > 250)

# Draw a boxplot of the mean versus Group without outliers
library(ggplot2)
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days 1-60")

```



In RATS dataset there are three group so two-sample t-test might not work. ANOVA test would be better. 


```{r}

library(dplyr)
library(tidyr)
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(baseline ~ Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)


```


Another way to do one-way ANOVA test. 

```{r}
res.aov <- aov(baseline ~ Group, data = RATSL8S2)

summary(res.aov)
```


As the p-value is less than the significance level 0.001 , we can conclude that there are significant differences between the groups. 

But with one-way anova the p-value only indicates that some of the group means are different, but we don’t know which pairs of groups are different.

We did not learn in the course which statistical test to use with data with three groups to see which of the groups are different but I found this one:

```{r}
TukeyHSD(res.aov)
```

According to Tukeys test the difference between groups 1 and 2 and between groups 1 and 3 is significant with adjusted p value close to 0. There is no significant difference between groups 2 and 3. 



*Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)*


In BPRS data 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated
on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks.



```{r}
library(dplyr)

BPRSL <- read.table("data/BPRSL.csv", header = T, sep=",")

# Factor treatment & subject
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)


# Glimpse the data

glimpse(BPRSL)

#structure of the data
str(BPRSL)

#dimensions of the data
dim(BPRSL)

#variable names of the data
names(BPRSL)

#view the data in table format
#view(BPRSL)

#summaries of the variables
summary(BPRSL)



```

Visualization of the BPRS data:


```{r}
# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

In the graphical illustration shows overlap between two treatment groups suggesting that there is no or only little difference between the two treatment groups.




Fitting the Independence Model to the BPRS data:








Regression model for the BPRS data

```{r}

# create a regression model BPRS_reg
BPRS_reg <-  lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```


Random intercept model:
Fitting a random intercept model allows the linear regression fit for each patient to differ in intercept from other patients.



```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

fit the random intercept and random slope model to the bprs data.

```{r}
# create a random intercept and random slope model
library(lme4)
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```


Fitting a random intercept and slope model that allows for a treatment group × time interaction.

```{r}
library(lme4)
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model

summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRSL with the observed bprs values
observed_plot <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = treatment)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)), name = "Observed bprs") +
  theme(legend.position = "top")+
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) 

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

library(dplyr)
library(tidyr)
# Create a new column fitted to RATSL
BPRSL <-  BPRSL %>% mutate(fitted = Fitted)

# draw the plot of BPRSL with the Fitted values of bprs

fitted_plot <- ggplot(BPRSL, aes(x = week, y = fitted, linetype = treatment)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)),name = "Fitted bprs") +
  theme(legend.position = "top")+
 scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both)

```
Visualization of the observed bprs and fitted bprs:

```{r}
library(ggpubr)

ggarrange(observed_plot,fitted_plot) 

```

